---
title: 'The tidyverse: Introduction to dplyr and tidyr'
author: "Jonathan Marshall"
output: html_document
---

```{r}
library(readxl)
library(dplyr)
library(tidyr)
# yay, this looks like everything, saved the values with zero in it. Let's do some checking though
all = read_excel('Roll-by-FYL-and-Ethnic-Group-2007-2017_fullsheet.xlsx', sheet="all_data") %>%
  filter(`Year: As at 1 July` == 2017)

# grab the schools
schools = all %>% select(`School: ID`, `School: Name`:`Region: MOE Local office`) %>% unique
schools = schools %>% rename_at(vars(starts_with("School: "), starts_with("Region: ")), function(x) { substring(x, 9)} ) %>%
  select(-ID, -`Education Region`, -`MOE Local office`) %>% 
  rename(School = Name, SchoolGender = Gender) %>%
  rename_all(funs(stringr::str_replace_all(., ' ', ''))) %>%
  arrange(School) %>%
  mutate(RegionalCouncil = stringr::str_replace(RegionalCouncil, " Region", ""),
         TerritorialAuthority = stringr::str_replace(TerritorialAuthority, " City", ""),
         TerritorialAuthority = stringr::str_replace(TerritorialAuthority, " District", ""),
         TerritorialAuthority = stringr::str_replace(TerritorialAuthority, "Auckland- ", "Auckland: "),
         KuraType = stringr::str_replace(KuraType, " \\(Section 15[5-6]\\)", ""),
         Decile = readr::parse_number(Decile))

# and now the student rolls
roll = all %>% select(starts_with('Student'), `School: ID`, `School: Name`) %>%
  rename_at(vars(starts_with("Student: ")), function(x) { substring(x, 10) } ) %>%
  rename(SchoolID = `School: ID`, School = `School: Name`,
         EthnicGroup = `Ethnic Group`, Students = `Students (∑ Values)`,
         Level = `Year level`) %>%
  select(-`Year level group`) %>%
  select(School, Gender, EthnicGroup, Level, Students) %>%
  arrange(School, Level) %>%
  mutate(EthnicGroup = case_when(EthnicGroup == "European\\Pākehā" ~ "European",
                                 TRUE ~ EthnicGroup))

# tidy up some fields


# Save for later
saveRDS(roll, 'data/roll_with_macrons.rds')
saveRDS(schools, 'data/schools_with_macrons.rds')

# Remove the macrons
roll_nm = roll %>% mutate(EthnicGroup = case_when(EthnicGroup == 'Māori' ~ "Maori", TRUE ~ EthnicGroup))
schools_nm = schools %>% mutate(KuraType = case_when(KuraType == 'Kura Kupapa Māori' ~ "Kura Kupapa Maori", TRUE ~ KuraType))

saveRDS(roll_nm, 'data/roll.rds')
saveRDS(schools_nm, 'data/schools.rds')
# TODO: Could potentially expand to include zeros, but no real point in doing so.
```

Things we can do:

## summarise

How many rows do we have in total?

How many students?

What is the mean, sd number of students per gender, ethnicity, school, year level and year?

```{r}
roll %>% summarise(Rows = n(), Total = sum(`Students (∑ Values)`))
roll %>% summarise(Rows = n(), Mean = mean(`Students (∑ Values)`), Min = min(`Students (∑ Values)`))
```

## arrange


- top_n()


## filter

rolls:

Find all Year 14/15 students.
Find all Students of Ethnicity <foo>
Find all male students in 2017

schools:

## select/rename

rolls:

Retrieve just the number of students, their gender, and ethnic groups.

## mutate

Create primary/secondary column.
Collapse ethnic groups?
Collapse last 3 years?
Convert ID to a factor.

## arrange

Find the entry with the most students.
Find the first school alphabetically.
Arrange the data by year level.




## group_by




```{r}
```